# Classificaiton

We are solving a problem of binary classification of motor imagery tasks (left/right fist) on per-trial EEG feautres (6 channels x ~15 tsfeatures). We will use multiple models to compare and discuss their performance and efficiency.

We start with loading the processed dataset.

```{r}
# Imports
library(tidyverse)
library(tsfeatures)
library(tidymodels)
library(discrim)

# install.packages("naivebayes") # Naive Bayes
# install.packages("kknn") # KNN
# install.packages("ranger") # Random Forest
```

```{r}
# Read the trail-level features file
features <- read_csv("data/processed/trial_features.csv")
str(features)
```

## Feature Distributions

We visualize the density distributions of the extracted features for the single electrode (`C3`), comparing the left and right classes to see the difference in distributions in the features between the two classes.

```{r}
features |>
  select(starts_with("C3_"), label) |>
  pivot_longer(cols = starts_with("C3_"), names_to = "feature", values_to = "value") |>
  ggplot(aes(x = value, fill = label, color = label)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ feature, scales = "free") +
  theme_minimal() +
  labs(
    title = "Density Distributions of C3 Features by Class",
    x = "Feature Value",
    y = "Density"
  )
```

All features are very similar. We quantify this by comparing the distributions of the `left` and `right` classes for each `C3` feature using the KS test.

```{r}
features |>
  select(starts_with("C3_"), label) |>
  pivot_longer(cols = starts_with("C3_"), names_to = "feature", values_to = "value") |>
  group_by(feature) |>
  summarise(
    ks_stat = ks.test(value[label == "left"], value[label == "right"])$statistic,
    p_value = ks.test(value[label == "left"], value[label == "right"])$p.value,
    .groups = "drop"
  ) |>
  arrange(p_value)
```

KS test shows significant difference being only in entropy (0.0628) and stability (0.0606) which is nonetheless still low. This suggests that individual features are not discriminative on their own.

## Preprocessing 

We prepare the dataset for further classification tasks keeping it identical for all models.

```{r}
# Factorize target
features <- features |>
  mutate(label = as.factor(label))

# Splitting
set.seed(42)
data_split <- group_initial_split(features, group = subject_id, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

# Base Recipe
# Predict label using everything else, but exclude ID-like columns from predictors
# We also scale features and remove highly correlated ones to stabilize models
base_recipe <- recipe(label ~ ., data = train_data) |>
  update_role(subject_id, trial_id, run, new_role = "ID") |>
  step_impute_mean(all_numeric_predictors()) |>
  step_zv(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_corr(all_numeric_predictors(), threshold = 0.9)
```

## Logistic Regression Baseline

We build a Logistic Regression model to serve as a baseline for other non-linear models.

```{r}
# Model Specification
logreg_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

# Workflow
logreg_wf <- workflow() |>
  add_recipe(base_recipe) |>
  add_model(logreg_spec)

# Fitting
logreg_fit <- logreg_wf |> fit(data = train_data)

# Evaluation
logreg_preds <- augment(logreg_fit, new_data = test_data)

# Print metrics (Accuracy, Sensitivity, Specificity, ROC AUC)
eval_metrics <- metric_set(accuracy, sensitivity, specificity, roc_auc)
eval_metrics(logreg_preds, truth = label, estimate = .pred_class, .pred_left)
# Confusion Matrix
logreg_preds |>
  conf_mat(truth = label, estimate = .pred_class)
```

As expected, logistic regression shows poor results, since the problem requires non-linear decision boundaries and complex interactions between the features.

## Naive Bayes Classifier

```{r}
# Model Specification
nb_spec <- naive_Bayes() |>
  set_engine("naivebayes") |>
  set_mode("classification")

# Workflow
nb_wf <- workflow() |>
  add_recipe(base_recipe) |>
  add_model(nb_spec)

# Fitting
nb_fit <- nb_wf |> fit(data = train_data)

# Evaluation
nb_preds <- augment(nb_fit, new_data = test_data)

# Print metrics
eval_metrics(nb_preds, truth = label, estimate = .pred_class, .pred_left)
# Confusion Matrix
nb_preds |>
  conf_mat(truth = label, estimate = .pred_class)
```

## KNN Classifier

```{r}
# Model Specification
knn_spec <- nearest_neighbor(neighbors = 10) |>
  set_engine("kknn") |>
  set_mode("classification")

# Workflow
knn_wf <- workflow() |>
  add_recipe(base_recipe) |>
  add_model(knn_spec)

# Fitting
knn_fit <- knn_wf |> fit(data = train_data)

# Evaluation
knn_preds <- augment(knn_fit, new_data = test_data)

# Print metrics
eval_metrics(knn_preds, truth = label, estimate = .pred_class, .pred_left)
# Confusion Matrix
knn_preds |>
  conf_mat(truth = label, estimate = .pred_class)
```

## Random Forest Classifier

```{r}
# Model Specification
rf_spec <- rand_forest(trees = 500) |>
  set_engine("ranger") |>
  set_mode("classification")

# Workflow
rf_wf <- workflow() |>
  add_recipe(base_recipe) |>
  add_model(rf_spec)

# Fitting
rf_fit <- rf_wf |> fit(data = train_data)

# Evaluation
rf_preds <- augment(rf_fit, new_data = test_data)

# Print metrics
eval_metrics(rf_preds, truth = label, estimate = .pred_class, .pred_left)
# Confusion Matrix
rf_preds |>
  conf_mat(truth = label, estimate = .pred_class)
```

## Combined Evaluation

```{r}
# Combined ROC Curve Plot
bind_rows(
  logreg_preds |> mutate(model = "Logistic Regression"),
  nb_preds |> mutate(model = "Naive Bayes"),
  knn_preds |> mutate(model = "KNN"),
  rf_preds |> mutate(model = "Random Forest")
) |>
  group_by(model) |>
  roc_curve(truth = label, .pred_left) |>
  autoplot() +
  labs(title = "ROC Curves: All Models Compared")
```

```{r}
# Print evaluation metrics and confusion matrices for all models
models <- list(
  "Logistic Regression" = logreg_preds,
  "Naive Bayes" = nb_preds,
  "KNN" = knn_preds,
  "Random Forest" = rf_preds
)

# Combined metrics table (metrics as columns)
metrics_tbl <- purrr::imap_dfr(models, ~ {
  res <- eval_metrics(.x, truth = label, estimate = .pred_class, .pred_left)
  res$model <- .y
  res
}) |> 
  select(model, .metric, .estimator, .estimate) |> 
  tidyr::pivot_wider(id_cols = model, names_from = .metric, values_from = .estimate, names_sep = "_")

print(metrics_tbl)
```