---
title: "EEG Stability Score Prediction — Linear Regression"
output: html_notebook
---

# 1. Setup & Data Loading

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(tidymodels)
library(glmnet)
library(corrplot)
library(vip)
```

We load the processed subject features and keep only the mean EEG features alongside the target variable `stability_score`. Subjects with any missing values are removed.

```{r}
subject_features <- read.csv("../data/processed/subject_features.csv")

reg_data <- subject_features |>
  select(subject_id, ends_with("_mean"), stability_score) |>
  drop_na()

cat("Subjects after cleaning:", nrow(reg_data), "\n")
cat("Number of predictors:", ncol(reg_data) - 2, "\n")
```

------------------------------------------------------------------------

# 2. Exploratory Data Analysis

## 2.1 Summary Statistics

```{r}
summary_df <- reg_data %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), list(
    min    = ~min(., na.rm = TRUE),
    q1     = ~quantile(., 0.25, na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    mean   = ~mean(., na.rm = TRUE),
    q3     = ~quantile(., 0.75, na.rm = TRUE),
    max    = ~max(., na.rm = TRUE),
    sd     = ~sd(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  separate(metric, into = c("feature", "stat"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = "stat", values_from = "value")

print(summary_df, n = 90, width = Inf)
```

## 2.2 Distribution of Features

Each predictor's distribution is inspected via histograms. Heavily skewed features may affect model assumptions.

```{r feature-distributions, fig.width=12, fig.height=10}
df_long <- reg_data %>%
  select(where(is.numeric), -subject_id) %>%
  pivot_longer(cols = everything(), names_to = "feature", values_to = "value")

ggplot(df_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~feature, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of EEG Mean Features and Stability Score",
       x = "Value", y = "Count")
```

## 2.3 Relationship Between Predictors and Stability Score

Scatter plots show the marginal linear relationship between each EEG feature and the target, providing a first justification for using linear regression.

```{r scatter-vs-target, fig.width=12, fig.height=10}
reg_data %>%
  select(where(is.numeric), -subject_id) %>%
  pivot_longer(-stability_score, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, y = stability_score)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linewidth = 0.8) +
  facet_wrap(~feature, scales = "free_x") +
  theme_minimal() +
  labs(title = "EEG Features vs. Stability Score",
       x = "Feature Value", y = "Stability Score")
```

## 2.4 Correlation Heatmap

The correlation matrix reveals multicollinearity among predictors, which motivates the use of regularized regression (Lasso) alongside standard OLS.

```{r correlation-heatmap, fig.width=10, fig.height=9}
cor_matrix <- reg_data %>%
  select(where(is.numeric), -subject_id) %>%
  cor(use = "complete.obs")

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.cex = 0.7,
         tl.col = "black",
         col = colorRampPalette(c("#2166ac", "white", "#d6604d"))(200),
         title = "Feature Correlation Matrix",
         mar = c(0, 0, 1.5, 0))
```

------------------------------------------------------------------------

# 3. Train / Test Split

We split the data 70/30 into training and test sets. All preprocessing parameters are estimated from the training set only to avoid data leakage.

```{r}
set.seed(42)
reg_data_split <- reg_data %>% initial_split(prop = 0.7)
train_data <- training(reg_data_split)
test_data  <- testing(reg_data_split)

x_train_raw <- train_data %>% select(-stability_score, -subject_id)
y_train_raw <- train_data %>% pull(stability_score)
x_test_raw  <- test_data  %>% select(-stability_score, -subject_id)
y_test_raw  <- test_data  %>% pull(stability_score)

cat("Training set:", nrow(train_data), "subjects\n")
cat("Test set:    ", nrow(test_data),  "subjects\n")
```

------------------------------------------------------------------------

# 4. Modelling

We compare four approaches to predict `stability_score` from EEG mean features.

-   **Model 1 — OLS (raw data):** included as a **cautionary baseline only**. With \~52 subjects and 84 predictors (p ≫ n), OLS is severely underdetermined: it will fit the training set perfectly while being meaningless on unseen data. This is not a valid model — it is shown precisely to illustrate *why* we need regularization.
-   **Model 2 — OLS (normalized):** the same underdetermined model after z-score normalization. Predictive performance is identical (normalization does not change OLS fit), but the scale-equalized coefficients make the overfitting pattern even more visible. Also kept as a demonstration, not a competing model.
-   **Model 3 — Lasso:** regularized linear regression fitted via **tidymodels**. The penalty λ is selected by 10-fold cross-validation. Lasso shrinks irrelevant coefficients to exactly zero, performing automatic feature selection and avoiding the overfit problem above.
-   **Model 4 — Random Forest:** a non-linear ensemble method included as the non-linear alternative to Lasso, fitted via **tidymodels** with default hyperparameters.

```{r}
rmse_fn <- function(actual, predicted) sqrt(mean((actual - predicted)^2, na.rm = TRUE))
r2_fn   <- function(actual, predicted) cor(actual, predicted)^2
```

```{r}
# Remove subject_id once — all model chunks use these data frames
train_model_data <- train_data %>% select(-subject_id)
test_model_data  <- test_data  %>% select(-subject_id)

# 10-fold CV folds for Lasso penalty tuning
set.seed(42)
cv_folds <- vfold_cv(train_model_data, v = 10)
```

## 4.1 Model 1: OLS on Raw Data

> **Demonstration only.** With more predictors than subjects (p ≫ n), OLS has infinitely many solutions and will memorize the training set. The purpose here is to show the overfit signature (train R² ≈ 1, test performance collapses) that motivates regularization.

```{r}
model_raw <- lm(y_train_raw ~ ., data = x_train_raw)

pred_raw_train <- predict(model_raw, newdata = x_train_raw)
pred_raw_test  <- predict(model_raw, newdata = x_test_raw)

metrics_raw <- data.frame(
  Model   = "OLS (Raw)",
  Dataset = c("Train", "Test"),
  RMSE    = c(rmse_fn(y_train_raw, pred_raw_train), rmse_fn(y_test_raw, pred_raw_test)),
  R2      = c(r2_fn(y_train_raw,   pred_raw_train), r2_fn(y_test_raw,   pred_raw_test))
)

print(metrics_raw)
```

## 4.2 Model 2: OLS on Normalized Data

> **Demonstration only.** Z-score normalization does not fix the p ≫ n problem — train and test metrics are identical to Model 1 (modulo numerical precision). The value here is pedagogical: with standardized coefficients, the magnitude of overfit is more legible.

Z-score normalization is applied using training set statistics only to prevent data leakage.

```{r}
train_means <- x_train_raw %>% summarise(across(everything(), ~mean(.x, na.rm = TRUE)))
train_sds   <- x_train_raw %>% summarise(across(everything(), ~sd(.x, na.rm = TRUE)))

normalize <- function(data, means, sds) {
  data %>% mutate(across(everything(), ~(.x - means[[cur_column()]]) / sds[[cur_column()]]))
}

x_train_norm <- normalize(x_train_raw, train_means, train_sds)
x_test_norm  <- normalize(x_test_raw,  train_means, train_sds)

# Remove zero-variance columns (would produce NaN after normalization)
zero_var <- names(x_train_norm)[sapply(x_train_norm, function(x) sd(x, na.rm = TRUE) == 0)]
if (length(zero_var) > 0) {
  cat("Removing zero-variance columns:", paste(zero_var, collapse = ", "), "\n")
  x_train_norm <- x_train_norm %>% select(-all_of(zero_var))
  x_test_norm  <- x_test_norm  %>% select(-all_of(zero_var))
}

model_norm <- lm(y_train_raw ~ ., data = x_train_norm)

pred_norm_train <- predict(model_norm, newdata = x_train_norm)
pred_norm_test  <- predict(model_norm, newdata = x_test_norm)

metrics_norm <- data.frame(
  Model   = "OLS (Normalized)",
  Dataset = c("Train", "Test"),
  RMSE    = c(rmse_fn(y_train_raw, pred_norm_train), rmse_fn(y_test_raw, pred_norm_test)),
  R2      = c(r2_fn(y_train_raw,   pred_norm_train), r2_fn(y_test_raw,   pred_norm_test))
)

print(metrics_norm)
```

## 4.3 Model 3: Lasso Regression

Lasso penalizes large coefficients and shrinks irrelevant predictors to exactly zero, solving the p ≫ n problem that breaks OLS. We use a **tidymodels** workflow: a recipe handles z-score normalization, and the penalty λ is selected via 10-fold cross-validation.

```{r}
# --- Recipe: normalize all predictors (fit on train only, applied to test) ---
lasso_recipe <- recipe(stability_score ~ ., data = train_model_data) %>%
  step_normalize(all_predictors())

# --- Model spec: Lasso (mixture = 1) with tunable penalty ---
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

# --- Workflow ---
lasso_wf <- workflow() %>%
  add_recipe(lasso_recipe) %>%
  add_model(lasso_spec)

# --- Tune: 50-point log-scale grid ---
lambda_grid <- grid_regular(penalty(), levels = 50)

set.seed(42)
lasso_tune <- tune_grid(
  lasso_wf,
  resamples = cv_folds,
  grid      = lambda_grid,
  metrics   = metric_set(rmse, rsq)
)

# --- Select best lambda by CV RMSE ---
best_lambda <- select_best(lasso_tune, metric = "rmse")
cat("Best lambda:", round(best_lambda$penalty, 6), "\n")

# --- Finalize workflow and fit on full training set ---
lasso_final <- finalize_workflow(lasso_wf, best_lambda)
lasso_fit   <- fit(lasso_final, data = train_model_data)

# --- Predict ---
pred_lasso_train <- predict(lasso_fit, new_data = train_model_data)$.pred
pred_lasso_test  <- predict(lasso_fit, new_data = test_model_data)$.pred

# --- Metrics ---
metrics_lasso <- data.frame(
  Model   = "Lasso",
  Dataset = c("Train", "Test"),
  RMSE    = c(rmse_fn(y_train_raw, pred_lasso_train), rmse_fn(y_test_raw, pred_lasso_test)),
  R2      = c(r2_fn(y_train_raw,   pred_lasso_train), r2_fn(y_test_raw,   pred_lasso_test))
)

print(metrics_lasso)

# --- Features selected ---
lasso_coef_tidy <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

n_selected <- sum(lasso_coef_tidy$estimate != 0)
cat("\nLasso selected", n_selected, "features out of", ncol(x_train_raw), "\n")
```

## 4.4 Model 4: Random Forest

Random Forest is the non-linear alternative to Lasso. It makes no linearity assumption and handles correlated predictors naturally. We use a **tidymodels** workflow with default hyperparameters.

```{r}
rf_spec <- rand_forest(mode = "regression") %>%
  set_engine("randomForest")

rf_wf <- workflow() %>%
  add_recipe(recipe(stability_score ~ ., data = train_model_data)) %>%
  add_model(rf_spec)

set.seed(42)
rf_fit <- fit(rf_wf, data = train_model_data)

pred_rf_train <- predict(rf_fit, new_data = train_model_data)$.pred
pred_rf_test  <- predict(rf_fit, new_data = test_model_data)$.pred

metrics_rf <- data.frame(
  Model   = "Random Forest",
  Dataset = c("Train", "Test"),
  RMSE    = c(rmse_fn(y_train_raw, pred_rf_train), rmse_fn(y_test_raw, pred_rf_test)),
  R2      = c(r2_fn(y_train_raw,   pred_rf_train), r2_fn(y_test_raw,   pred_rf_test))
)

print(metrics_rf)
```

------------------------------------------------------------------------

# 5. Model Comparison

## 5.1 Unified Metrics Table

The OLS models are included for reference only. The meaningful comparison is Lasso vs. Random Forest on the test set.

```{r}
all_metrics <- rbind(metrics_raw, metrics_norm, metrics_lasso, metrics_rf)
all_metrics <- all_metrics %>% mutate(across(where(is.numeric), ~round(., 4)))
print(all_metrics)
```

## 5.2 Predicted vs. Actual (Lasso & Random Forest — Test Set)

Points close to the diagonal indicate accurate predictions.

```{r pred-vs-actual, fig.width=10, fig.height=5}
bind_rows(
  data.frame(actual = y_test_raw, predicted = pred_lasso_test, Model = "Lasso"),
  data.frame(actual = y_test_raw, predicted = pred_rf_test,    Model = "Random Forest")
) %>%
  ggplot(aes(x = actual, y = predicted, color = Model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  facet_wrap(~Model) +
  scale_color_manual(values = c("Lasso" = "steelblue", "Random Forest" = "darkorange")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Predicted vs. Actual Stability Score (Test Set)",
       x = "Actual Stability Score",
       y = "Predicted Stability Score")
```

------------------------------------------------------------------------

# 6. Model Diagnostics (Lasso)

Regression diagnostics are run on the **Lasso model**, which is the actual linear model of interest. Running these on the OLS normalized model would be uninformative: with p ≫ n, OLS perfectly interpolates the training set, leaving residuals of essentially zero.

## 6.1 Residuals vs. Fitted

Residuals should be randomly scattered around zero with no systematic pattern. A funnel shape would indicate heteroscedasticity; a curve would suggest the linear model is missing non-linear structure.

```{r residuals-vs-fitted, fig.width=7, fig.height=5}
resid_df <- data.frame(
  fitted    = pred_lasso_train,
  residuals = y_train_raw - pred_lasso_train
)

ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, color = "firebrick", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "orange", linewidth = 0.8) +
  theme_minimal() +
  labs(title = "Residuals vs. Fitted Values (Lasso — Train Set)",
       x = "Fitted Values", y = "Residuals")
```

## 6.2 Q-Q Plot of Residuals

If residuals are approximately normally distributed, the points should lie closely along the diagonal. Systematic departures suggest the error distribution is non-Gaussian, which can affect inference even though it does not directly affect Lasso predictions.

```{r qq-plot, fig.width=6, fig.height=5}
ggplot(resid_df, aes(sample = residuals)) +
  stat_qq(color = "steelblue", alpha = 0.7) +
  stat_qq_line(color = "firebrick", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Q-Q Plot of Residuals (Lasso — Train Set)",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

------------------------------------------------------------------------

# 7. Feature Importance — Lasso Selected Features

Lasso shrinks irrelevant coefficients to exactly zero. The plot below shows only the features retained by the model, ranked by their effect on `stability_score`.

```{r lasso-coef-plot, fig.width=8, fig.height=6}
lasso_coef_df <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)", estimate != 0) %>%
  arrange(desc(abs(estimate)))

ggplot(lasso_coef_df, aes(x = reorder(term, estimate), y = estimate,
                           fill = estimate > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick"),
                    labels = c("TRUE" = "Positive", "FALSE" = "Negative")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Lasso Selected Features and Their Coefficients",
       x = "Feature", y = "Coefficient Estimate", fill = "Direction")
```

------------------------------------------------------------------------

# 8. Feature Importance — Random Forest

Random Forest does not produce coefficients, but it records how much each predictor reduces impurity across all trees. The `vip` package extracts these importance scores from the fitted engine.

```{r rf-importance, fig.width=8, fig.height=6}
rf_fit |>
  extract_fit_engine() |>
  vip(num_features = 20, aesthetics = list(fill = "steelblue", color = "white")) +
  theme_minimal() +
  labs(title = "Random Forest — Top 20 Feature Importances",
       x = "Feature", y = "Importance (Mean Decrease in Node Impurity)")
```

## 8.1 Lasso vs. Random Forest: Feature Agreement

```{r}
# Lasso selected features
lasso_selected <- lasso_coef_df %>%
  transmute(
    Feature = term,
    Lasso_Coef = round(estimate, 4)
  )

n_lasso <- nrow(lasso_selected)

# RF top-N importance
rf_vi <- rf_fit %>%
  extract_fit_engine() %>%
  vip::vi() %>%
  transmute(
    Feature = Variable,
    RF_Rank = row_number()
  )

# Join and compare
comparison_df <- lasso_selected %>%
  left_join(rf_vi, by = "Feature") %>%
  mutate(
    RF_Agrees = ifelse(
      !is.na(RF_Rank) & RF_Rank <= n_lasso,
      "YES",
      "NO"
    )
  ) %>%
  arrange(desc(abs(Lasso_Coef)))

cat(sprintf(
  "Lasso kept %d features | RF agrees on %d (top-%d overlap)\n\n",
  n_lasso,
  sum(comparison_df$RF_Agrees == "Yes"),
  n_lasso
))

print(comparison_df)
```

------------------------------------------------------------------------

# 9. Summary and Conclusion

```{r}
test_metrics <- all_metrics %>%
  filter(Dataset == "Test") %>%
  filter(Model == "Random Forest" | Model == "Lasso")%>%
  arrange(RMSE) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

print(test_metrics)
```

The two OLS models are not meaningful competitors — with p ≫ n they memorize the training set perfectly and fail on unseen data. The real comparison is **Lasso vs. Random Forest**. Lasso provides a sparse linear model with automatic feature selection; Random Forest captures non-linear structure and interactions without any explicit selection step.

The features where both models agree — where Lasso assigns a non-zero coefficient *and* RF ranks the same feature in its top-N — are the strongest candidates for a genuine EEG stability signature. That convergence across two completely different selection mechanisms (coefficient shrinkage vs. node impurity) is hard to explain by noise alone. Features that Lasso kept but RF ranks lower likely have a clean linear relationship with stability that gets diluted inside the ensemble; features RF ranks highly that Lasso dropped point to non-linear effects or interactions a penalized linear model cannot represent. If RF outperforms Lasso on the test set, those disagreements are the most likely explanation.
